---
title: "Assignment Report on LULC Mapping using Available Satellite
  RS Data and VGI In-situ Data in R"
author: "Krishna Dev Singh"
date: "2023-11-05"
output:
  html_document: default
  editor_options: null
  markdown:
    wrap: 77
  pdf_document: default
  word_document: default
---
```{r include=FALSE}
library(raster)
library(tidyverse)
library(sf)
library(rpart)
library(rpart.plot)
library(rasterVis)
library(mapedit)
library(mapview)
library(caret)
library(e1071)

setwd("D:/R_Project")
# Load the Landsat image and the training data
band1 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B1.TIF")
band2 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B2.TIF")
band3 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B3.TIF")
band4 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B4.TIF")
band5 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B5.TIF")
band6 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B6.TIF")
band7 <- raster("Satellite_Image/LC08_L2SP_141041_20211226_20211230_02_T1_SR_B7.TIF")
# Read the individual bands and stack them
img <- stack(band1, band2, band3, band4, band5, band6, band7)
names(img) <- c(paste0("B", 1:7, coll = ""))
# Define the extent
aoi <- st_read("shp_files/Sarlahi.shp")
aoi <-st_transform(aoi, crs(img))
# Clip the stacked raster to the custom extent
image <- crop(img, aoi)

# plot the clipped raster True Color Composite
par(col.axis="white",col.lab="white",tck=0)
plotRGB(image, r = 4, g = 3, b = 2, axes = TRUE, 
        stretch = "lin", main = "True Color Composite")
box(col="white")

# plot the clipped raster False Color Composite
par(col.axis="white",col.lab="white",tck=0)
plotRGB(image, r = 5, g = 4, b = 3, axes = TRUE, stretch = "lin", main = "False Color Composite")
box(col="white")

#We calculate NDVI using the following equation (NIR - Red)/(NIR + Red). The [[ ]] notation specifies the bands, in this case band 5 for NIR and band 4 for Red, within the multi-band raster.
ndvi <- (image[[5]] - image[[4]])/(image[[5]] + image[[4]])

as(ndvi, "SpatialPixelsDataFrame") %>% 
  as.data.frame() %>%
  ggplot(data = .) +
  geom_tile(aes(x = x, y = y, fill = layer)) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "NDVI for Sarlahi", 
       x = " ", 
       y = " ") +
  scale_fill_gradient(high = "#CEE50E", 
                      low = "#087F28",
                      name = "NDVI")

# Read training points
training_shp <- st_read("shp_files/LULC_Crop-Types_In-SituData2021_USP.shp")
training_shp <-st_transform(training_shp, crs(img))
LULC_aoi <- st_intersection(training_shp, aoi)

# Use mutate to create the "id" column based on the mapping
class_names <- c("Agriculture"= 1, "Commercial"= 2, "CULARCH"= 3, "Forest"= 4, "PublicUse"= 5, "Residential"= 6, "WaterBodies"= 7)
LULC_aoi <- LULC_aoi %>%
  mutate(id = class_names[data_LULC])
write_sf(LULC_aoi, "shp_files/Sarlahi_TP1.shp")
training <- st_read("shp_files/Sarlahi_TP1.shp", quiet = TRUE)

#Plot the training points
ggplot() +
  geom_sf(data = training, aes(color = data_LULC ), size = 0.5) +
  scale_color_manual(values = c('cyan', 'burlywood', 'darkgreen', 'blue','grey','red','yellow')) +
  labs(title = "Classification points by land use") +
  theme(panel.background = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())

# Split the AOI into training and testing datasets (70% training, 30% testing)
set.seed(17)  # For reproducibility
splitIndex <- createDataPartition(training$data_LULC, p = 0.7, 
                                  list = FALSE, 
                                  groups = 1)
training_points <- training[splitIndex, ]
testing_points <- training[-splitIndex, ]
unique(testing_points$data_LULC)
#Extracting spectral values from the raster
training_points <- as(training_points, 'Spatial')
df <- raster::extract(image, training_points) %>%
  round()
head(df)

#Classifying the imagery

df <- data.frame(training_points$data_LULC, df)

model.class <- rpart(as.factor(training_points.data_LULC)~., data = df, method = 'class', minsplit = 7)
#plot the decision tree.
rpart.plot(model.class, box.palette = 0, main = "Classification Tree")

#Using the model, we predict on the entire image and set the classes to the four land cover categories.

pr <- predict(image, model.class, type ='class', progress = 'text') %>% 
  ratify()


levels(pr) <- levels(pr)[[1]] %>% 
  mutate(legend = c("Agriculture", "Commercial","Forest","Residential","Waterbodies"))

levelplot(pr, maxpixels = 1e6,
          col.regions = c('cyan', 'burlywood', 'grey', 'darkgreen','red','yellow','blue'),
          scales=list(draw=FALSE),
          main = " Classified Imagery")

predicted <- raster::extract(pr, testing_points) %>% 
  as.data.frame() %>% 
  rename(id = ".")

df1 <- data.frame(
  obs_id = testing_points$id,
  obs = testing_points$data_LULC,
  pred = predicted$id
)
# Replace numeric values with character labels
df1$pred <- ifelse(df1$pred == 1, "Agriculture",
                 ifelse(df1$pred == 2, "Commercial",
                        ifelse(df1$pred == 3, "CULARCH",
                               ifelse(df1$pred == 4, "Forest",
                                      ifelse(df1$pred == 5, "PublicUse",
                                             ifelse(df1$pred == 6, "Residential",
                                                    ifelse(df1$pred== 7, "WaterBodies", NA)))))))
actual_classes <- df1$obs
confusion_matrix <- table(Predicted = df1$pred, Actual = actual_classes)
print(confusion_matrix)


# Convert the confusion matrix to a data frame
confusion_matrix_df <- as.data.frame(as.table(confusion_matrix))

# Rename the columns for better labels
colnames(confusion_matrix_df) <- c("Predicted", "Actual", "Frequency")

# Create the heatmap using ggplot2
heatmap_plot <- ggplot(confusion_matrix_df, aes(Actual, Predicted, fill = Frequency)) +
  geom_tile() +
  geom_text(aes(label = Frequency), vjust = 1, size = 4) +
  scale_fill_gradient(low = "pink", high = "red") +  # Define color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted")

# Display the heatmap
print(heatmap_plot)

# Assuming your dataframe is named 'data'
df1$obs <- as.factor(df1$obs)
df1$pred <- as.factor(df1$pred)

# Align the levels of 'pred' with the levels of 'obs'
levels(df1$pred) <- levels(df1$obs)

# Calculate confusion matrix, precision, recall, F-score, accuracy, and Kappa
confusion_matrix <- confusionMatrix(df1$pred, df1$obs)
print("Confusion Matrix:")
print(confusion_matrix)
```

# Executive Summary

In this report,In this report, I had generated LULC Map of Sarlahi District
using LandSat 8 image and training data obtained from secondary source 
volunteered geographic information(VGI) in R. We have used R, along with various 
packages, for geospatial analysis and machine learning to achieve the objectives. 
I would like to thank Prof. Arun Pratihasta who inspired and guided through the
entire project.

**Key Findings:**

1.  **Data Preparation**: I loaded in-situ data and have several
    proprocessing tasks such as cleaning, removing unusable column, and
    merging columns as required to represent land use and crop types.

2.  **Satellite Data**: I had downloaded Landsat 8 image from USGS 
    Earth Explorer and processed in R.

3.  **NDVI Analysis**: I had calculated the NDVI values from the satellite
    data.

4.  **Machine Learning**: I had implemented a rpart classifier model to predict
    land cover classes .The model was trained and tested using in-situ data in the
    ratio of 70%:30%.

5.  **Model Evaluation**: We evaluated the model's performance using a
    confusion matrix and calculated accuracy, precision, recall, and
    F1-score for each land cover class. We also determined producer and
    user accuracy values.

**Recommendations:**

Based on the project findings, the following recommendations can be
made:

-   The more insitu data should be collect with more accuratly for
    refining the predictive model.

-   Multiple indices should be used with addition of NDVI to have better
    predictions.

-   Consider conducting field surveys and data collection to further
    validate the model's predictions and enhance its reliability.

-   Extend the analysis to a larger geographical area to assess land use
    and land cover patterns more widely

-   Explore with using more machine learning techniques to compare the
    results to have better reliability.

# Table of Contents

-   [Introduction](#introduction)
-   [Methods](#methods)
-   [Results](#results)
-   [Discussion](#discussion)
-   [Conclusion](#conclusion)
-   [References](#references)

# Introduction {#introduction}



Land use and land cover mapping using Landsat 8 images in R is a vital technique
in environmental science, urban planning, and resource management. Landsat 8, a 
satellite equipped with multispectral sensors, captures data in various spectral
bands, allowing for the classification of land features. In R, researchers can
harness powerful geospatial packages like 'raster' and 'rgdal' to process and
analyze Landsat 8 imagery. By employing spectral indices, such as NDVI, and
machine learning algorithms like random forests, these tools enable accurate 
classification of land cover types, including forests, agriculture, urban areas,
and water bodies. The resulting maps offer valuable insights for sustainable
land management and conservation efforts.

**Context**

This report focuses on a specific case study involving the mapping of
land use land cover types using available satellite remote sensing data and 
VGI data in the context of Madesh Province of Nepal specially Sarlahi District. 
The in-situ data has been collected after the end of Mansoon in the month of
October to Janurary. The sample size is approx 600, and is backed with
photographic proof. The study area has been chosen due to its significance in 
agriculture and the presence of diverse crop types. The regions is known 
epicenter of for crop production for the country.

### Objectives

The primary objectives of this report are as follows:

-   To analyze the feasibility and effectiveness of classifying LULC
    types using satellite remote sensing data and VGI.

-   To visualize the insitu data over the map

-   To assess the accuracy and performance of the classification model
    with confusion matrix, overall accuracy and individual class
    performance.



# Methods

## ***Data Collection:***

**Satellite Remote Sensing Data:** The first component of our data
collection involved obtaining satellite remote sensing data. We acquired
multi-spectral Landsat 8 imagery fromvUSGS Earth explorervcovering the study
area during the month of December, 2021. The metada of Image is as follows:
GROUP = LANDSAT_METADATA_FILE
  GROUP = PRODUCT_CONTENTS
    ORIGIN = "Image courtesy of the U.S. Geological Survey"
    DIGITAL_OBJECT_IDENTIFIER = "https://doi.org/10.5066/P9OGBGM6"
    LANDSAT_PRODUCT_ID = "LC08_L2SP_141041_20211226_20211230_02_T1"
    PROCESSING_LEVEL = "L2SP"
    COLLECTION_NUMBER = 02
    COLLECTION_CATEGORY = "T1"
    OUTPUT_FORMAT = "GEOTIFF"
    MAP_PROJECTION = "UTM"
    DATUM = "WGS84"
    ELLIPSOID = "WGS84"
    UTM_ZONE = 45
    GRID_CELL_SIZE_PANCHROMATIC = 15.00
    GRID_CELL_SIZE_REFLECTIVE = 30.00
    GRID_CELL_SIZE_THERMAL = 30.00
    ORIENTATION = "NORTH_UP"
    RESAMPLING_OPTION = "CUBIC_CONVOLUTION"
```{r }
# plot the clipped raster True Color Composite
par(col.axis="white",col.lab="white",tck=0)
plotRGB(image, r = 4, g = 3, b = 2, axes = TRUE, 
        stretch = "lin", main = "True Color Composite")
box(col="white")
```

```{r }
# plot the clipped raster False Color Composite
par(col.axis="white",col.lab="white",tck=0)
plotRGB(image, r = 5, g = 4, b = 3, axes = TRUE, stretch = "lin", main = "False Color Composite")
box(col="white")
```


**Volunteered Geographic Information (VGI):** In parallel with satellite
data, we had availed in-situ VGI data. The in-situ data used was
collected after the end of the monsoon from October 23, 2021, to January
6, 2021. Participants provided valuable ground-level information about
crop types in the study are with photograph and geographical coordinates
associated with each observation.


## Analysis 
In this section, we will describe the methods and techniques
used to process and analyze the collected remote sensing and in-situ
data. Our primary goal is to classify land use and crop types in the
study area by integrating satellite imagery with in-situ data.

**Libraries:**

-   **R Language:** We conducted our data analysis and geospatial
    processing using the R programming language. R is a powerful tool
    for statistical analysis and geospatial data handling, making it
    well-suited for our research objectives.

-   **R Packages:** Several R packages were essential for our analysis.

    -   **sf, terra and raster:** These packages allowed us to work with
        geospatial data, including shapefiles and raster imagery. We
        calculated the Normalized Difference Vegetation Index (NDVI)
        using these packages to assess vegetation health.

    -   **dplyr:** This package was used for data manipulation, allowing
        us to filter, mutate, and aggregate datasets effectively.

    -   **viridis:** We utilized the viridis package for defining custom
        color palettes in our visualizations, making it easier to
        distinguish between land cover classes.

    -   **randomForest:** We employed the random forest algorithm for
        supervised classification tasks.

    -   **knitr,tidyr,ggplot2:** These packages were used for data
        visualization, including the creation of plots, table and
        heatmaps.

    -   **leaflet,leaflet.extras:** For the interactive mapping
        component of our analysis, we utilized the leaflet package.

### ***Data Preprocessing:*** 

The data was clean, check for unusable columns and removed it. The new
column LU_CT was created to store crop types and other build up area
merges as Mixed_area.

**Data Projection and Alignment:** To ensure data compatibility, the
satellite remote sensing data and VGI were projected and aligned to a
common coordinate reference system (CRS, EPSG:32645).

**NDVI Calculation:** The NDVI was computed using the red and
near-infrared bands of the satellite imagery. The NDVI calculation was
performed using the formula (NIR - Red) / (NIR + Red), where NIR
represents the near-infrared reflectance values and Red represents the
red band reflectance values. This index is very helpful for
distinguishing vegetation health and land use. \### NDVI Visualitzation

```{r }
#We calculate NDVI using the following equation (NIR - Red)/(NIR + Red). The [[ ]] notation specifies the bands, in this case band 5 for NIR and band 4 for Red, within the multi-band raster.
ndvi <- (image[[5]] - image[[4]])/(image[[5]] + image[[4]])

as(ndvi, "SpatialPixelsDataFrame") %>% 
  as.data.frame() %>%
  ggplot(data = .) +
  geom_tile(aes(x = x, y = y, fill = layer)) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "NDVI for Sarlahi", 
       x = " ", 
       y = " ") +
  scale_fill_gradient(high = "#CEE50E", 
                      low = "#087F28",
                      name = "NDVI")
```

#### I had choosen area of interest as Sarlai District.


```{r}
# Read training points
training_shp <- st_read("shp_files/LULC_Crop-Types_In-SituData2021_USP.shp")
training_shp <-st_transform(training_shp, crs(img))
LULC_aoi <- st_intersection(training_shp, aoi)

# Use mutate to create the "id" column based on the mapping
class_names <- c("Agriculture"= 1, "Commercial"= 2, "CULARCH"= 3, "Forest"= 4, "PublicUse"= 5, "Residential"= 6, "WaterBodies"= 7)
LULC_aoi <- LULC_aoi %>%
  mutate(id = class_names[data_LULC])
write_sf(LULC_aoi, "shp_files/Sarlahi_TP1.shp")
training <- st_read("shp_files/Sarlahi_TP1.shp", quiet = TRUE)
#Plot the training points
ggplot() +
  geom_sf(data = training, aes(color = data_LULC ), size = 0.5) +
  scale_color_manual(values = c('cyan', 'burlywood', 'darkgreen', 'blue','grey','red','yellow')) +
  labs(title = "Classification points by land use") +
  theme(panel.background = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())

```
**Data Splitting** The dataset was clipped with the shapefiles of sarlahi district
and was further divided into training and testing sets in a proportion of 70% 
for training and 30% for testing. 
```{r }
# Split the AOI into training and testing datasets (70% training, 30% testing)
set.seed(17)  # For reproducibility
splitIndex <- createDataPartition(training$data_LULC, p = 0.7, 
                                  list = FALSE, 
                                  groups = 1)
training_points <- training[splitIndex, ]
testing_points <- training[-splitIndex, ]
unique(testing_points$data_LULC)
```

**Classification:**

We use the rpart function in R to create a classification tree model. Here's an explanation of the code:
model.class <- rpart(...): This line of code assigns the result of the rpart function to the object model.class. The model.class will contain the trained classification tree model.
as.factor(training_points.data_LULC)~.: This part of the code specifies the formula for the model. It suggests that you are trying to predict the factor variable training_points.data_LULC based on the values of all other variables in the df dataset. In other words, you're building a classification tree to predict the Land Use and Land Cover (LULC) class based on the other variables in the dataset.
data = df: This specifies the dataset df from which the variables for the model are taken.
method = 'class': The method parameter is set to 'class', indicating that you are creating a classification tree, as opposed to a regression tree.
minsplit = 7: The minsplit parameter determines the minimum number of data points required to split a node. In this case, a node will only be split if it has at least 7 data points. This parameter helps control the complexity of the tree.
```{r }
#Extracting spectral values from the raster
training_points <- as(training_points, 'Spatial')
df <- raster::extract(image, training_points) %>%
  round()
head(df)
```

## **Classification Results:**

The primary objective of this study was to classify land use land cover
types using the integrated remote sensing and in-situ data. 
```{r }
setwd("D:/R_Project")
#Classifying the imagery
df <- data.frame(training_points$data_LULC, df)
model.class <- rpart(as.factor(training_points.data_LULC)~., data = df, method = 'class', minsplit = 7)
#plot the decision tree.
rpart.plot(model.class, box.palette = 0, main = "Classification Tree")
```
```{r }
#Using the model, we predict on the entire image and set the classes to the four land cover categories.

pr <- predict(image, model.class, type ='class', progress = 'text') %>% 
  ratify()

levels(pr) <- levels(pr)[[1]] %>% 
  mutate(legend = c("Agriculture", "Commercial","Forest","Residential","Waterbodies"))

levelplot(pr, maxpixels = 1e6,
          col.regions = c('cyan', 'burlywood', 'grey', 'darkgreen','red','yellow','blue'),
          scales=list(draw=FALSE),
          main = " Classified Imagery")
```
**Overall Accuracy:** The overall accuracy of the classification model
was approximately 69.23%.
**Heatmap Visualization of Confusion Matrix:**

To evaluate the performance of our classification model, we created a
heatmap of the confusion matrix. This heatmap visually represents the
accuracy and misclassification of different land use and crop types.

```{r }
# Create the heatmap using ggplot2
heatmap_plot <- ggplot(confusion_matrix_df, aes(Actual, Predicted, fill = Frequency)) +
  geom_tile() +
  geom_text(aes(label = Frequency), vjust = 1, size = 4) +
  scale_fill_gradient(low = "pink", high = "red") +  # Define color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted")

# Display the heatmap
print(heatmap_plot)
```
**Confusion Matrix:** The confusion matrix below provides detailed
information on the model's performance. It displays the number of
correctly and incorrectly classified samples for each class. It shows
how well our model correctly classified different land types and helps
us calculate important performance metrics like accuracy, precision,
recall, and F1-score. This matrix enables us to assess both the model's
overall performance and its ability to correctly identify specific land
use types.

```{r echo=FALSE}
# Calculate confusion matrix, precision, recall, F-score, accuracy, and Kappa
confusion_matrix <- confusionMatrix(df1$pred, df1$obs)
print("Confusion Matrix:")
print(confusion_matrix)
```

# Conclusion {#conclusion}

This project highlights the effectiveness of using Rpart machine
learning models for land use and land cover classification. It
demonstrates that integrating in-situ data with satellite imagery can
significantly enhance classification accuracy. The model exhibited
nice results and also identifies areas for future planning. By continuing
to refine the model and incorporating more extensive datasets, we can
further explore the potential of remote sensing and machine learning for
applications in environmental and urban areas. This work lays the
foundation for future research and applications in land use and land
cover classification.



# References {#references}

1. Volunteered In-Situ Data for Agriculture Crop Mapping: A Case Study of
Nepal By U. S. Panday , A. K. Pratihast , J. Aryal , R. Kayastha 

2. Pratihast, A. K., DeVries, B., Avitabile, V., de
Bruin, S., Kooistra, L., Tekle, M., &
Herold, M. (2014). Combining satellite data
and community-based observations for
forest monitoring. Forests.
https://doi.org/10.3390/f5102464

3. Pratihast, A. K., Herold, M., Avitabile, V., de
Bruin, S., Bartholomeus, H., Souza, C. M.,
& Ribbe, L. (2013). Mobile devices for
community-based REDD+ monitoring: A
case study for central Vietnam. Sensors
(Switzerland), 13, 21–38.
https://doi.org/10.3390/s130100021
